---
title: "Final CFB Prediction Model"
author: "Nolan Santacroce"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)

library(cfbfastR)
library(caret)
library(dplyr)
library(ggplot2)
library(randomForest)
library(tidyr)
library(zoo)

source("cfb_model_rebuild.R")

if (Sys.getenv("CFBD_API_KEY") == "") {
  message("Set CFBD_API_KEY in .Renviron before running CFBD rating pulls.")
}
```

## Model Configuration

This version intentionally removes raw `home` and `away` team names from the model inputs. The model only sees current/as-of team quality differentials, coach-context differentials, home-field/neutral-site context, and optional experimental feature groups.

```{r config}
MODEL_YEARS <- 2014:2025
TRAIN_YEARS <- 2014:2024
PREDICT_YEAR <- 2025

# Default excludes PFF because it can overrate some G5 teams and is noisy across seasons.
PFF_GROUP <- "none"

# Season-level injury totals are useful for playoff context but leak information for weekly ATS.
# Keep FALSE for weekly/backtest work until a week-specific injury file is available.
INCLUDE_SEASON_INJURIES <- FALSE
PRESEASON_INPUT_FILE <- "preseason_team_inputs.csv"

FIRST_BACKTEST_YEAR <- 2018
MAX_WEEK <- 18
```

## Data Loading

```{r load-data}
pbp_all <- cfbfastR::load_cfb_pbp(MODEL_YEARS)

sp_ratings_all <- fetch_sp_ratings(MODEL_YEARS)
elo_ratings_all <- fetch_weekly_elo_ratings(MODEL_YEARS, weeks = 1:MAX_WEEK)

coach_raw <- read_coach_data("coach_ratings.csv")
pff_raw <- read_pff_data("pff_team_data.csv")
preseason_raw <- read_preseason_team_inputs(PRESEASON_INPUT_FILE)
injury_raw <- read_injury_data("final_for_sure_1.csv")
game_lines <- load_game_lines("game_lines.csv")
```

## Coach Rating Audit

The coach file is cleaned to one row per `school-year`, with the primary coach selected by games coached. Ratings are based only on prior coach history, so a coach's current-season performance is not leaked into games from that season.

```{r coach-audit}
coach_duplicate_keys <- coach_raw %>%
  count(school, year, name = "rows") %>%
  filter(rows > 1) %>%
  arrange(desc(rows), school, year)

coach_clean <- build_clean_coach_features(coach_raw)

coach_duplicate_keys %>% head(20)

coach_clean %>%
  filter(team %in% c("Alabama", "Clemson", "Indiana", "SMU")) %>%
  arrange(team, year) %>%
  select(
    team, year, coach_name, coach_rating_pre,
    coach_career_win_pct_before, coach_srs_mean_before,
    coach_recent3_srs_before, coach_school_tenure_before,
    midseason_coach_change
  )
```

## Build Modeling Table

Every model row is one game. For each team, the feature snapshot is the latest information available before that game's week. The final feature set is made of home-minus-away differentials, so old brand strength is not directly learned through team names.

Week 0/1 games are tagged as `preseason`, Weeks 2-4 as `early_season`, and Week 5+ as `in_season`. Current-season form features are zero-weighted for Week 0/1, partially weighted in Weeks 2-4, and fully weighted once the season has enough data.

```{r build-model-table}
model_table <- build_game_model_table(
  pbp = pbp_all,
  sp_ratings = sp_ratings_all,
  elo_ratings = elo_ratings_all,
  coach_raw = coach_raw,
  pff_raw = pff_raw,
  preseason_raw = preseason_raw,
  injury_raw = injury_raw,
  lines = game_lines,
  pff_group = PFF_GROUP,
  include_season_injuries = INCLUDE_SEASON_INJURIES,
  max_week = MAX_WEEK
)

feature_cols <- default_feature_columns(
  model_table,
  pff_group = PFF_GROUP,
  include_season_injuries = INCLUDE_SEASON_INJURIES
)

feature_cols

model_table %>%
  summarise(
    games = n(),
    seasons = n_distinct(year),
    features = length(feature_cols),
    preseason_games = sum(game_phase == "preseason"),
    early_season_games = sum(game_phase == "early_season"),
    in_season_games = sum(game_phase == "in_season"),
    ats_games_with_lines = sum(!is.na(close_spread_home)),
    preseason_input_rows = nrow(preseason_raw)
  )
```

## Week 0/1 Preseason Model

The preseason model layer uses prior-season/team-history features, cleaned coach context, optional preseason priors, and the light manual preseason file. It does not use current-season rolling EPA/WPA for Week 0/1.

```{r preseason-input-contract}
names(read_preseason_team_inputs(PRESEASON_INPUT_FILE))

preseason_manual_match_cols <- intersect(
  c("projected_qb_tier_diff", "preseason_power_adjustment_diff"),
  names(model_table)
)

preseason_model_rows <- model_table %>% filter(game_phase == "preseason")
preseason_manual_rows_matched <- if (length(preseason_manual_match_cols) == 0) {
  0
} else {
  sum(rowSums(!is.na(preseason_model_rows[, preseason_manual_match_cols, drop = FALSE])) > 0)
}

tibble(
  games = nrow(preseason_model_rows),
  current_form_features_used = sum(abs(coalesce(preseason_model_rows$off_epa_last3_phase_weighted, 0)) > 0),
  preseason_manual_rows_matched = preseason_manual_rows_matched
)
```

```{r week01-backtest}
week01_backtest_comparison <- compare_week01_models(
  model_table,
  first_eval_year = FIRST_BACKTEST_YEAR,
  last_eval_year = max(MODEL_YEARS),
  pff_group = PFF_GROUP
)

week01_backtest_comparison
```

## Rolling Backtest

The backtest trains on all prior seasons, applies recency-based row weighting, then predicts the next season. This exposes whether the model is adapting to current team quality rather than carrying too much historical Alabama/Clemson style brand weight.

```{r rolling-backtest}
rolling_backtest <- run_rolling_backtest(
  model_table,
  first_eval_year = FIRST_BACKTEST_YEAR,
  last_eval_year = max(MODEL_YEARS),
  min_train_year = min(MODEL_YEARS),
  feature_cols = feature_cols
)

rolling_backtest$metrics
```

```{r rolling-backtest-plot}
rolling_backtest$metrics %>%
  ggplot(aes(x = eval_year, y = su_accuracy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Rolling Straight-Up Accuracy by Evaluation Season",
    x = "Evaluation season",
    y = "Straight-up accuracy"
  ) +
  theme_minimal()
```

## Final Training and Diagnostics

```{r final-model}
final_model <- train_margin_model(
  model_table,
  train_years = TRAIN_YEARS,
  feature_cols = feature_cols,
  half_life_years = 3,
  max_multiplier = 6
)

final_model$model

season_predictions <- model_table %>%
  filter(year == PREDICT_YEAR) %>%
  predict_model_table(final_model, .)

evaluate_predictions(season_predictions)
diagnostic_slices(season_predictions)
```

```{r variable-importance}
varImp(final_model$model)
```

## PFF Experiments

Run this chunk only when you want to test whether PFF improves out-of-sample results. PFF is excluded from the default model unless one of these groups improves rolling playoff and ATS metrics.

```{r pff-experiments, eval=FALSE}
pff_group_results <- evaluate_pff_groups(
  pbp = pbp_all,
  sp_ratings = sp_ratings_all,
  elo_ratings = elo_ratings_all,
  coach_raw = coach_raw,
  pff_raw = pff_raw,
  preseason_raw = preseason_raw,
  injury_raw = injury_raw,
  lines = game_lines,
  groups = c("none", "overall", "unit", "trenches", "pass_rush"),
  first_eval_year = FIRST_BACKTEST_YEAR
)

pff_group_results
```

## Matchup Prediction

The matchup helper uses as-of snapshots from the rebuilt model table. `spread_home` follows the standard convention where a home favorite is negative, for example `-6.5`.

```{r matchup-example, eval=FALSE}
predict_matchup(
  home_team = "Indiana",
  away_team = "Clemson",
  season = 2025,
  week = 16,
  model_bundle = final_model,
  model_table = model_table,
  neutral_site = TRUE,
  spread_home = NA_real_
)
```

## Weekly Article Prep

Use `predict_week()` for Week 0/1 or any weekly card. The output is model-only: margin, winner, ATS edge, confidence, and top feature drivers.

```{r predict-week-example, eval=FALSE}
week_card <- tibble(
  season = c(2025, 2025),
  week = c(1, 1),
  home = c("Indiana", "Alabama"),
  away = c("Clemson", "SMU"),
  neutral_site = c(FALSE, TRUE),
  open_spread_home = c(NA_real_, NA_real_),
  close_spread_home = c(NA_real_, NA_real_)
)

predict_week(
  schedule = week_card,
  model_bundle = final_model,
  model_table = model_table
)
```

## Betting Lines Contract

Add closing/opening spreads to `game_lines.csv` when available. `close_spread_home` should be negative when the home team is favored and positive when the home team is an underdog.

Required columns:

```{r line-columns, echo=FALSE}
names(load_game_lines("game_lines.csv"))
```

## Preseason Input Contract

Fill `preseason_team_inputs.csv` once before the season. Suggested scale: `projected_qb_tier` from 1-5 where 5 is elite, continuity fields as 0/1, and `preseason_power_adjustment` as your manual points adjustment relative to the model's baseline.

The lower-maintenance workflow is to paste/download public source tables into the source CSVs, then generate `preseason_team_inputs.csv`. Your own entries should go in `preseason_manual_overrides.csv`, which is applied last.

Source files:

- `preseason_source_returning_starters.csv`: returning starters/snap percentages
- `preseason_source_portal.csv`: transfer portal team rank/score
- `preseason_source_staff.csv`: current and prior head coach/coordinator names
- `preseason_manual_overrides.csv`: QB tier, power adjustment, and any continuity fixes

```{r automated-preseason-inputs, eval=FALSE}
automated_preseason <- build_automated_preseason_inputs(
  season = 2026,
  team_context = attr(model_table, "model_components")$team_snapshots,
  output_path = PRESEASON_INPUT_FILE
)

automated_preseason %>%
  arrange(returning_starter_value_rank) %>%
  select(
    team, season, returning_starter_value_rank, returning_starter_value,
    returning_starters_total, quality_context_score, portal_net_rating,
    projected_qb_tier, preseason_power_adjustment
  ) %>%
  head(25)
```
